Attribution of undesirable character traits, rather than trait-based dehumanization, predicts punishment decisions

### 2024-07-08 manuscript.Rmd
Update manuscript.Rmd
Updated to match accepted manuscript

We revisit the hypothesized causal relationship between trait-based dehumanization and harm in the context of endorsing harsh punishment for criminals. In Studies 1A and 1B, we seek to conceptually replicate the key findings of Bastian and colleagues [-@Bastian2013], suggesting that the extent to which participants animalistically (Study 1A) and mechanistically (Study 1B) dehumanize criminals predicts the severity of the punishment participants endorse for them. In Studies 2A and 2B, we adopt a similar design but incorporate socially undesirable traits into our stimulus set. This addition to the design allows us to investigate whether trait-based dehumanization, undesirable trait attribution, or both predict the severity of punishment. Following Bastian and colleagues, and to understand the generalisability of our findings, we investigate these questions in relation to two different types of crime (violent crime and theft). In Studies 3A and 3B, we seek to investigate a similar question using an experimental design. We present participants with vignettes in which criminals are described using character traits that differ in how socially desirable they are and whether or not they are unique to humans. We also measure how these varying descriptions influence participants' parole decisions. This allows us to directly measure whether there is a causal relationship between trait-based dehumanization and punishment, independent of ingroup preference.

@@

All studies received ethical approval from the Psychology Departmental Ethics Committee at the University of York (approval number 926). All data collection occurred online, and the studies were created and administered using Qualtrics (https://www.qualtrics.com). Participants were recruited through the online platform Prolific (https://www.prolific.co), with an independent sample recruited for each study. Informed consent was obtained at the start of each session according to approved ethical guidelines. Inclusion criteria for each study included adult participants fluent in English who had never been to prison for committing a crime and had a Prolific approval rating of at least 90% (95% for Studies 3A and 3B). Increases in Prolific's recommended rate of compensation for participation during data collection meant the reward ranged from approximately £7 per hour in Studies 1A and 1B to approximately £8 in the other four studies. Assumption testing and analyses were conducted using *SPSS* and *RStudio*. All studies were pre-registered on AsPredicted.com before commencing data collection. Links to pre-registration documents, data files [@Brennan2024], a fully computationally reproducible version of the manuscript, and electronic supplementary materials including the stimuli used for each study can be found at: http://doi.org/10.17605/OSF.IO/D4CVP.

@@

```{r figure1, fig.cap="Results of Study 1. Seemingly in line with Bastian and colleagues (2013), greater animalistic (left) and mechanistic (right) dehumanization of violent criminals (Study 1A, top) and thieves (Study 1B, bottom) was associated with harsher punishment.", fig.align="center", echo=FALSE}

@@

Study 1B investigates whether the relationship found in Study 1A replicates when participants are asked to judge a different type of criminal activity. In Study 1B, we examined whether animalistic and mechanistic forms of dehumanization, as operationalized by @Bastian2013, are associated with the harshness of punishment endorsed for individuals who commit theft. The design, materials and analysis plan were similar to that used in Study 1A, except that the scenarios involved theft rather than violent crime. In investigating a different type of crime, we follow the example set by Bastian et al. and seek to understand the generalisability of our results.

@@

Based on the same power analysis used in Study 1A, a sample of 100 participants was collected, with `r gendertable1b$Freq[2]+1` identifying as male and `r gendertable1b$Freq[1]` as female. Ages ranged from `r minage1b ` to `r maxage1b` (*M* = `r meanage1b `, *SD* = `r round(sdage1b, digits = 2) `). The attention checks in Study 1A were also used in Study 1B. Ten participants failed one or both attention checks, and their data was omitted and replaced as per our pre-registration. Participation took an average of nine minutes.

@@

The measures of dehumanization and punishment were identical to those used in Study 1A.

**Vignettes** All participants responded to the same five vignettes, each detailing a crime involving theft (see electronic supplementary materials). As in Study 1A, an effort was made to ensure the vignettes were similar in structure and amount of detail. Once again, all of the perpetrators had gender-neutral names. An example of one of the theft vignettes is as follows: "*Until their recent arrest, Charlie had worked as a till operator at a local charity shop supporting individuals experiencing homelessness. Charlie had been stealing cash amounts varying from £5 to £50 from the tills almost daily over a five-year period. Police revealed that Charlie had stolen several thousand pounds from the charity shop while working there*."

@@

A power analysis using G\*Power indicated that a sample size of `r ceiling((Exp2APA$v)+(Exp2APA$u)+1)` would allow us to detect a medium effect size ($f^2$ = `r Exp2APA$f2`), with three predictors (trait attribution; trait desirability; attribution*desirability), an alpha of `r Exp2APA$sig.level` and power of `r Exp2APA$power`. To counterbalance the sample equally and allow for the exclusion of outliers, a sample of `r length(Study2A_numeric$CaseID)` was collected. Within the sample, `r genderTable2A$Freq[1]` identified as female, `r genderTable2A$Freq[2]` as male, and `r genderTable2A$Freq[3]` as non-binary. Ages ranged from `r min(Study2A_numeric$Age)` to `r max(Study2A_numeric$Age)` ($M$ = `r round (mean(Study2A_numeric$Age), digits = 2)`, $SD$ = `r round (sd(Study2A_numeric$Age), digits = 2)`). Similar to Studies 1A and 1B, two attention checks were included in this study. As per our pre-registered plan, 16 participants failed one or both attention checks; thus, their data were omitted and replaced. Participation took an average of nearly eight minutes.

@@

**Vignettes**. All participants read the same vignettes describing violent crimes as in Study 1A.

@@

**Harshness of punishment endorsed**. The same single-item scale for measuring the harshness of punishment endorsed in Studies 1A and 1B was employed in Study 2A.

@@

The procedure in Study 2A mirrored that of Studies 1A and 1B.

@@

The moderated regression showed no significant effects of human nature trait attribution on punishment, $b$ = `r round(exp2AHNBeta$estimate[2], digits = 2)`  [`r round(exp2AHNBeta$conf.low[2], digits = 2)`,  `r round(exp2AHNBeta$conf.high[2], digits = 2)`], $t$ = `r round(exp2AHNBeta$statistic[2], digits = 2)`, $p$ = `r format.pval(exp2AHNBeta$p.value[2], digits = 2, eps = 0.001, nsmall = 3)`. The effect reported by Bastian and colleagues [-@Bastian2013], whereby mechanistic dehumanization predicted harsher punishment endorsement, which we replicated in Studies 1A and 1B, did not appear when undesirable human nature traits were included in our measures.

@@

```{r figure2, fig.cap="Results of Study 2: The relationship between trait attribution and punishment for violent criminals (Study 2A, top) and thieves (Study 2B, bottom) depends on the social desirability of the traits.", fig.align="center", echo=FALSE}

@@

Study 2B sought to replicate the results of Study 2A but with thieves as the target group rather than violent criminals. We examined whether the apparent relationship between trait-based dehumanization and the endorsement of harsh punishment for thieves is better explained by the desirability of the traits incorporated into the stimulus set. We investigate this question using a very similar design and procedure to Study 2A, with the exception that the vignettes are those used in Study 1B detailing crimes involving theft. As in Study 2A, we hypothesise that trait desirability will moderate the relationship between human trait attribution and punishment. More specifically, we predict the extent to which criminals are denied socially desirable character traits, and attributed socially undesirable character traits, will predict endorsement of harsher punishment.

@@

The power analysis described in Study 2A informed the sample size for Study 2B. A separate sample of `r length(Study2B_numeric$CaseID)` participants was collected, of whom `r genderTable2B$Freq[2]` identified as male, `r genderTable2B$Freq[1]` as female, and `r spell_out(genderTable2B$Freq[3])` as non-binary. Ages ranged from `r min(Study2B_numeric$Age)` to `r max(Study2B_numeric$Age)` (*M*= `r round(mean(Study2B_numeric$Age),digits=1)`, *SD* = `r round(sd(Study2B_numeric$Age),digits=2)`). Data submitted by 20 participants who did not pass one or both checks were omitted and replaced. Participation took an average of nine and a half minutes.

@@

This study utilized a mixed-methods design, matching that of Study 2A. The same attention checks used in Studies 1A, 1B and 2A were used in Study 2B.

@@

**Vignettes**. All participants responded to the same five vignettes used in Study 1B detailing scenarios involving criminals committing theft.

**Trait attribution**. The same scales for measuring animalistic dehumanization, mechanistic dehumanization, and punishment used in Study 2A were used in Study 2B.

@@

The procedure in Study 2B was identical to that of Study 2A, except for the vignettes describing crimes involving theft rather than violence.

@@

Eight highly influential cases were omitted from the analysis (remaining sample *N* = `r length(UH_harm_rmCooks_data$CaseID)`). A moderated regression tested for a relationship between average scores of uniquely human traits and harshness of punishment endorsed and whether this interacted with trait desirability. The moderated regression showed no significant effects of uniquely human trait attribution on punishment *b* = `r round(exp2BUHBeta$estimate[2], digits = 2)`  [`r round(exp2BUHBeta$conf.low[2], digits = 2)`, `r round(exp2BUHBeta$conf.high[2], digits = 2)`], *t* = `r round(exp2BUHBeta$statistic[2], digits = 2)`, *p* = `r format.pval(exp2BUHBeta$p.value[2], digits = 3)`. Replicating the results of Study 2A, when socially undesirable traits were incorporated into the stimulus set, there was no longer any relationship between trait-based dehumanization and punishment.

The interaction between uniquely human trait attribution and trait desirability was significant, *b* = `r round(exp2BUHBeta$estimate[4], digits = 2)` [`r round(exp2BUHBeta$conf.low[4], digits = 2)`, `r round(exp2BUHBeta$conf.high[4], digits = 2)`], *t* = `r round(exp2BUHBeta$statistic[4], digits = 2)`, *p* = `r format.pval(exp2BUHBeta$p.value[4], digits = 2, eps = 0.001, nsmall = 3)`. As illustrated in Figure \@ref(fig:figure2), the more participants attributed socially desirable traits to criminals, the less harshly participants felt they should be punished, *b* = `r round(exp2BUHSlope$simslopes$slopes$Est.[1], digits = 2)`, [`r round(exp2BUHSlope$simslopes$slopes$"2.5%"[1], digits = 2)`, `r round(exp2BUHSlope$simslopes$slopes$"97.5%"[1], digits = 2)`], *t* = `r round(exp2BUHSlope$simslopes$slopes$"t val."[1], digits = 2)`, *p* = `r format.pval(exp2BUHSlope$simslopes$slopes$"p"[1], digits = 2, eps = 0.001, nsmall = 3)`. The more participants attributed undesirable traits to criminals, the more harshly participants felt they should be punished, *b* = `r round(exp2BUHSlope$simslopes$slopes$Est.[2], digits = 2)`, [`r round(exp2BUHSlope$simslopes$slopes$"2.5%"[2], digits = 2)`, `r round(exp2BUHSlope$simslopes$slopes$"97.5%"[2], digits = 2)`], *t* = `r round(exp2BUHSlope$simslopes$slopes$"t val."[2], digits = 2)`, *p* = `r format.pval(exp2BUHSlope$simslopes$slopes$"p"[2], digits = 2, eps = 0.001, nsmall = 3)`. The model explained about `r round((exp2BUHR2$r.squared * 100), digits = 0)`% of the variance in endorsed harshness of punishment scores, $R^2$ = `r round(exp2BUHR2$r.squared, digits = 2)`, $F$(`r round(exp2BUHR2$fstatistic[2], digits = 0)`, `r round(exp2BUHR2$fstatistic[3], digits = 0)`) = `r round(exp2BUHR2$fstatistic[1], digits = 2)`.

@@

The analysis omitted six highly influential cases (remaining sample *N* = `r length(HN_harm_rmCooks_data$CaseID)`). A regression tested for a relationship between human trait attribution and punishment and whether this interacted with trait desirability. As in Study 2A, and contradicting the findings of Bastian and colleagues [-@Bastian2013], the moderated regression showed no significant relationship between human trait attribution and punishment, *b* = `r round(exp2BHNBeta$estimate[2], digits = 2)`  [`r round(exp2BHNBeta$conf.low[2], digits = 2)`,  `r round(exp2BHNBeta$conf.high[2], digits = 2)`], *t* = `r round(exp2BHNBeta$statistic[2], digits = 2)`, *p* = `r format.pval(exp2BHNBeta$p.value[2], digits = 3)`.

However, the interaction between human nature trait attribution and trait desirability was significant, *b* = `r round(exp2BHNBeta$estimate[4], digits = 2)`  [`r round(exp2BHNBeta$conf.low[4], digits = 2)`,  `r round(exp2BHNBeta$conf.high[4], digits = 2)`], *t* = `r round(exp2BHNBeta$statistic[4], digits = 2)`, *p* = `r format.pval(exp2BHNBeta$p.value[4], digits = 2, eps = 0.001, nsmall = 3)`. As can be seen in Figure \@ref(fig:figure2), simple slopes showed that the more participants attributed socially desirable human traits to criminals, the less harshly participants thought they should be punished, *b* = `r round(exp2BHNSlope$simslopes$slopes$Est.[1], digits = 2)`, [`r round(exp2BHNSlope$simslopes$slopes$"2.5%"[1], digits = 2)`, `r round(exp2BHNSlope$simslopes$slopes$"97.5%"[1], digits = 2)`], *t* = `r round(exp2BHNSlope$simslopes$slopes$"t val."[1], digits = 2)`, *p* = `r format.pval(exp2BHNSlope$simslopes$slopes$"p"[1], digits = 1)`. The more participants attributed socially undesirable human traits to criminals, the more harshly participants thought they should be punished, *b* = `r round(exp2BHNSlope$simslopes$slopes$Est.[2], digits = 2)`, [`r round(exp2BHNSlope$simslopes$slopes$"2.5%"[2], digits = 2)`, `r round(exp2BHNSlope$simslopes$slopes$"97.5%"[2], digits = 2)`], *t* = `r round(exp2BHNSlope$simslopes$slopes$"t val."[2], digits = 2)`, *p* = `r format.pval(exp2BHNSlope$simslopes$slopes$"p"[2], digits = 2, eps = 0.001, nsmall = 3)`. The model explained about `r round((exp2BHNR2$r.squared * 100), digits = 0)`% of the variance in endorsed harshness of punishment scores, $R^2$ = `r round(exp2BHNR2$r.squared, digits = 2)`, $F$(`r round(exp2BHNR2$fstatistic[2], digits = 0)`, `r round(exp2BHNR2$fstatistic[3], digits = 0)`) = `r round(exp2BHNR2$fstatistic[1], digits = 2)`. These data suggest that the apparent relationships between animalistic and mechanistic dehumanization and punishment reported in previous research [@Bastian2013] are better explained by the social desirability of the traits.

@@

In Study 3A, we used an experimental design to examine further the hypothesized causal relationship between trait-based dehumanization and punishment when controlling for the social desirability of human traits incorporated into the stimuli. We described criminals with traits that varied in desirability and perceived humanness creating a $2 \times 2$ design. We then measured participants' willingness to endorse parole for each criminal described. The dual model predicts that criminals who are described in uniquely human terms will be more likely to be granted parole. We predicted that criminals described in socially desirable terms will be more likely to be granted parole. In principle, this design allows us to detect independent effects of dehumanization and trait sociability or an interaction between the two. In Study 3A, we specifically measure the extent to which animalistic dehumanization is causally related to parole decisions. Thus, we included uniquely human traits and those shared with other animals in our measures. We predict that participants will be more likely to endorse parole for criminals described with socially desirable traits, regardless of whether or not those traits are uniquely human or shared with other animals.

@@

A power analysis using G\*Power, with effect size specification as in SPSS, indicated that a sample size of 135 would allow us to detect a medium effect size ($\eta_p^2$ = .09) with a $2 \times 2$ factorial, repeated measures design, an alpha of .05, and a power of .95. To counterbalance the sample equally, a sample of 136 participants was collected, of whom `r gen_data3A[1]` identified as female, `r gen_data3A[2]` as male, `r gen_data3A[3]` as non-binary, and `r gen_data3A[4]` who preferred not to indicate their gender identity.  Ages ranged from `r min_age3A` to `r max_age3A` (*M* = `r round(age_mean3A, digits = 1)`, *SD* = `r round(age_sd3A, digits = 2)`).  All participants were adults fluent in English who had never been to prison for committing a crime.  Due to a noticeable increase in failed attention checks during pilot data collection, the minimal approval rating on Prolific was raised from 90% to 95%.  Despite this, data from 46 participants were omitted and replaced due to failed attention checks. Three participants were mistakenly recruited after the intended sample size had been met, and thus, their data were excluded from analyses^[Including data submitted by excess participants in analyses yielded the same results as those reported.]. Participation took an average of seven minutes.

@@

**Vignettes**. All participants responded to the same four vignettes, each detailing a different scenario in which a criminal's eligibility for parole was assessed. Efforts were made to ensure that all four vignettes were similar in length, degree of detail, and contextual aspects, such as how long the criminal had spent in prison and who was described as attributing the traits to the criminal. In each vignette, the criminal's name and pronouns were gender-neutral, their age and ethnicity were not indicated, and their crime and sentence were not specified. The four vignettes can be seen in the Supplementary materials. In the uniquely human socially desirable condition, the criminal was described as *cultured, civilized, sophisticated*, and *moral*, while in the uniquely human condition socially undesirable, the criminal was described as *corrupt, controlling, arrogant* and *bitter*. In the animalistic desirable condition, the criminal was described as *energetic, trusting, genuine* and *having curiosity*, while in the animalistic undesirable condition, the criminal was described as *uncultured, unrefined, unsophisticated*, and *stupid*.

@@

**Agreement with parole**. The dependent variable, agreement with granting parole, was measured using the following single-item measure: '*I think (Alex/Sam/Robin/Jamie) should be granted parole*'. This measure appeared after each vignette, and participants indicated their agreement using an unmarked sliding scale ranging from Strongly Disagree (0) to Strongly Agree (100). The slider's initial starting point was always centred at 50.

@@

```{r figure3, fig.cap="Results of Studies 3A and 3B.  Criminals described with undesirable traits were less likely to be granted parole than were those described with desirable traits, regardless of whether or not those traits were uniquely human.", fig.align="center", out.width='90%', echo=FALSE}

@@

@@

The agreement with granting parole scale and attention check were the same as those used in Study 3A.

>>>>

We revisit the hypothesized causal relationship between trait-based dehumanization and harm in the context of endorsing harsh punishment for criminals. In studies 1A and 1B, we seek to conceptually replicate the key findings of Bastian and colleagues [-@Bastian2013], suggesting that the extent to which participants animalistically (study 1A) and mechanistically (study 1B) dehumanize criminals predicts the severity of the punishment participants endorse for them. In studies 2A and 2B, we adopt a similar design but incorporate socially undesirable traits into our stimulus set. This addition to the design allows us to investigate whether trait-based dehumanization, undesirable trait attribution, or both predict the severity of punishment. Following Bastian and colleagues, and to understand the generalisability of our findings, we investigate these questions in relation to two different types of crime (violent crime and theft). In studies 3A and 3B, we seek to investigate a similar question using an experimental design. We present participants with vignettes in which criminals are described using character traits that differ in how socially desirable they are and whether or not they are unique to humans. We also measure how these varying descriptions influence participants' parole decisions. This allows us to directly measure whether there is a causal relationship between trait-based dehumanization and punishment, independent of ingroup preference.

@@

All studies received ethical approval from the Psychology Departmental Ethics Committee at the University of York (approval number 926). All data collection occurred online, and the studies were created and administered using Qualtrics (https://www.qualtrics.com). Participants were recruited through the online platform Prolific (https://www.prolific.co), with an independent sample recruited for each study. Informed consent was obtained at the start of each session according to approved ethical guidelines. Inclusion criteria for each study included adult participants fluent in English who had never been to prison for committing a crime and had a Prolific approval rating of at least 90% (95% for studies 3A and 3B). Increases in Prolific's recommended rate of compensation for participation during data collection meant the reward ranged from approximately £7 per hour in studies 1A and 1B to approximately £8 in the other four studies. Assumption testing and analyses were conducted using *SPSS* and *RStudio*. All studies were pre-registered on AsPredicted.com before commencing data collection. Links to pre-registration documents, data files [@Brennan2024], a fully computationally reproducible version of the manuscript, and electronic supplementary materials including the stimuli used for each study can be found at: http://doi.org/10.17605/OSF.IO/D4CVP.

@@

```{r figure1, fig.cap="Results of study 1. Seemingly in line with Bastian and colleagues (2013), greater animalistic (left) and mechanistic (right) dehumanization of violent criminals (study 1A, top) and thieves (study 1B, bottom) was associated with harsher punishment.", fig.align="center", echo=FALSE}

@@

Study 1B investigates whether the relationship found in study 1A replicates when participants are asked to judge a different type of criminal activity. In study 1B, we examined whether animalistic and mechanistic forms of dehumanization, as operationalized by @Bastian2013, are associated with the harshness of punishment endorsed for individuals who commit theft. The design, materials and analysis plan were similar to that used in study 1A, except that the scenarios involved theft rather than violent crime. In investigating a different type of crime, we follow the example set by Bastian et al. and seek to understand the generalisability of our results.

@@

Based on the same power analysis used in study 1A, a sample of 100 participants was collected, with `r gendertable1b$Freq[2]+1` identifying as male and `r gendertable1b$Freq[1]` as female. Ages ranged from `r minage1b ` to `r maxage1b` (*M* = `r meanage1b `, *SD* = `r round(sdage1b, digits = 2) `). The attention checks in study 1A were also used in study 1B. Ten participants failed one or both attention checks, and their data was omitted and replaced as per our pre-registration. Participation took an average of nine minutes.

@@

The measures of dehumanization and punishment were identical to those used in study 1A.

**Vignettes** All participants responded to the same five vignettes, each detailing a crime involving theft (see electronic supplementary materials). As in study 1A, an effort was made to ensure the vignettes were similar in structure and amount of detail. Once again, all of the perpetrators had gender-neutral names. An example of one of the theft vignettes is as follows: "*Until their recent arrest, Charlie had worked as a till operator at a local charity shop supporting individuals experiencing homelessness. Charlie had been stealing cash amounts varying from £5 to £50 from the tills almost daily over a five-year period. Police revealed that Charlie had stolen several thousand pounds from the charity shop while working there*."

@@

A power analysis using G\*Power indicated that a sample size of `r ceiling((Exp2APA$v)+(Exp2APA$u)+1)` would allow us to detect a medium effect size ($f^2$ = `r Exp2APA$f2`), with three predictors (trait attribution; trait desirability; attribution*desirability), an alpha of `r Exp2APA$sig.level` and power of `r Exp2APA$power`. To counterbalance the sample equally and allow for the exclusion of outliers, a sample of `r length(Study2A_numeric$CaseID)` was collected. Within the sample, `r genderTable2A$Freq[1]` identified as female, `r genderTable2A$Freq[2]` as male, and `r genderTable2A$Freq[3]` as non-binary. Ages ranged from `r min(Study2A_numeric$Age)` to `r max(Study2A_numeric$Age)` ($M$ = `r round (mean(Study2A_numeric$Age), digits = 2)`, $SD$ = `r round (sd(Study2A_numeric$Age), digits = 2)`). Similar to studies 1A and 1B, two attention checks were included in this study. As per our pre-registered plan, 16 participants failed one or both attention checks; thus, their data were omitted and replaced. Participation took an average of nearly eight minutes.

@@

**Vignettes**. All participants read the same vignettes describing violent crimes as in study 1A.

@@

**Harshness of punishment endorsed**. The same single-item scale for measuring the harshness of punishment endorsed in studies 1A and 1B was employed in study 2A.

@@

The procedure in study 2A mirrored that of studies 1A and 1B.

@@

The moderated regression showed no significant effects of human nature trait attribution on punishment, $b$ = `r round(exp2AHNBeta$estimate[2], digits = 2)`  [`r round(exp2AHNBeta$conf.low[2], digits = 2)`,  `r round(exp2AHNBeta$conf.high[2], digits = 2)`], $t$ = `r round(exp2AHNBeta$statistic[2], digits = 2)`, $p$ = `r format.pval(exp2AHNBeta$p.value[2], digits = 2, eps = 0.001, nsmall = 3)`. The effect reported by Bastian and colleagues [-@Bastian2013], whereby mechanistic dehumanization predicted harsher punishment endorsement, which we replicated in studies 1A and 1B, did not appear when undesirable human nature traits were included in our measures.

@@

```{r figure2, fig.cap="Results of study 2: The relationship between trait attribution and punishment for violent criminals (study 2A, top) and thieves (study 2B, bottom) depends on the social desirability of the traits.", fig.align="center", echo=FALSE}

@@

Study 2B sought to replicate the results of study 2A but with thieves as the target group rather than violent criminals. We examined whether the apparent relationship between trait-based dehumanization and the endorsement of harsh punishment for thieves is better explained by the desirability of the traits incorporated into the stimulus set. We investigate this question using a very similar design and procedure to study 2A, with the exception that the vignettes are those used in study 1B detailing crimes involving theft. As in study 2A, we hypothesise that trait desirability will moderate the relationship between human trait attribution and punishment. More specifically, we predict the extent to which criminals are denied socially desirable character traits, and attributed socially undesirable character traits, will predict endorsement of harsher punishment.

@@

The power analysis described in study 2A informed the sample size for study 2B. A separate sample of `r length(Study2B_numeric$CaseID)` participants was collected, of whom `r genderTable2B$Freq[2]` identified as male, `r genderTable2B$Freq[1]` as female, and `r spell_out(genderTable2B$Freq[3])` as non-binary. Ages ranged from `r min(Study2B_numeric$Age)` to `r max(Study2B_numeric$Age)` (*M*= `r round(mean(Study2B_numeric$Age),digits=1)`, *SD* = `r round(sd(Study2B_numeric$Age),digits=2)`). Data submitted by 20 participants who did not pass one or both checks were omitted and replaced. Participation took an average of nine and a half minutes.

@@

This study used a mixed-methods design, matching that of study 2A. The same attention checks used in studies 1A, 1B and 2A were used in study 2B.

@@

**Vignettes**. All participants responded to the same five vignettes used in study 1B detailing scenarios involving criminals committing theft.

**Trait attribution**. The same scales for measuring animalistic dehumanization, mechanistic dehumanization, and punishment used in study 2A were used in study 2B.

@@

The procedure in study 2B was identical to that of study 2A, except for the vignettes describing crimes involving theft rather than violence.

@@

Eight highly influential cases were omitted from the analysis (remaining sample *N* = `r length(UH_harm_rmCooks_data$CaseID)`). A moderated regression tested for a relationship between average scores of uniquely human traits and harshness of punishment endorsed and whether this interacted with trait desirability. The moderated regression showed no significant effects of uniquely human trait attribution on punishment *b* = `r round(exp2BUHBeta$estimate[2], digits = 2)`  [`r round(exp2BUHBeta$conf.low[2], digits = 2)`, `r round(exp2BUHBeta$conf.high[2], digits = 2)`], *t* = `r round(exp2BUHBeta$statistic[2], digits = 2)`, *p* = `r format.pval(exp2BUHBeta$p.value[2], digits = 3)`. Replicating the results of study 2A, when socially undesirable traits were incorporated into the stimulus set, there was no longer any relationship between trait-based dehumanization and punishment.

The interaction between uniquely human trait attribution and trait desirability was significant, *b* = `r round(exp2BUHBeta$estimate[4], digits = 2)` [`r round(exp2BUHBeta$conf.low[4], digits = 2)`, `r round(exp2BUHBeta$conf.high[4], digits = 2)`], *t* = `r round(exp2BUHBeta$statistic[4], digits = 2)`, *p* `r format.pval(exp2BUHBeta$p.value[4], digits = 2, eps = 0.001, nsmall = 3)`. As illustrated in Figure \@ref(fig:figure2), the more participants attributed socially desirable traits to criminals, the less harshly participants felt they should be punished, *b* = `r round(exp2BUHSlope$simslopes$slopes$Est.[1], digits = 2)`, [`r round(exp2BUHSlope$simslopes$slopes$"2.5%"[1], digits = 2)`, `r round(exp2BUHSlope$simslopes$slopes$"97.5%"[1], digits = 2)`], *t* = `r round(exp2BUHSlope$simslopes$slopes$"t val."[1], digits = 2)`, *p* `r format.pval(exp2BUHSlope$simslopes$slopes$"p"[1], digits = 2, eps = 0.001, nsmall = 3)`. The more participants attributed undesirable traits to criminals, the more harshly participants felt they should be punished, *b* = `r round(exp2BUHSlope$simslopes$slopes$Est.[2], digits = 2)`, [`r round(exp2BUHSlope$simslopes$slopes$"2.5%"[2], digits = 2)`, `r round(exp2BUHSlope$simslopes$slopes$"97.5%"[2], digits = 2)`], *t* = `r round(exp2BUHSlope$simslopes$slopes$"t val."[2], digits = 2)`, *p* `r format.pval(exp2BUHSlope$simslopes$slopes$"p"[2], digits = 2, eps = 0.001, nsmall = 3)`. The model explained about `r round((exp2BUHR2$r.squared * 100), digits = 0)`% of the variance in endorsed harshness of punishment scores, $R^2$ = `r round(exp2BUHR2$r.squared, digits = 2)`, $F$(`r round(exp2BUHR2$fstatistic[2], digits = 0)`, `r round(exp2BUHR2$fstatistic[3], digits = 0)`) = `r round(exp2BUHR2$fstatistic[1], digits = 2)`.

@@

The analysis omitted six highly influential cases (remaining sample *N* = `r length(HN_harm_rmCooks_data$CaseID)`). A moderated regression tested for a relationship between human trait attribution and punishment and whether this interacted with trait desirability. As in study 2A, and contradicting the findings of Bastian and colleagues [-@Bastian2013], the moderated regression showed no significant relationship between human trait attribution and punishment, *b* = `r round(exp2BHNBeta$estimate[2], digits = 2)`  [`r round(exp2BHNBeta$conf.low[2], digits = 2)`,  `r round(exp2BHNBeta$conf.high[2], digits = 2)`], *t* = `r round(exp2BHNBeta$statistic[2], digits = 2)`, *p* = `r format.pval(exp2BHNBeta$p.value[2], digits = 3)`.

However, the interaction between human nature trait attribution and trait desirability was significant, *b* = `r round(exp2BHNBeta$estimate[4], digits = 2)`  [`r round(exp2BHNBeta$conf.low[4], digits = 2)`,  `r round(exp2BHNBeta$conf.high[4], digits = 2)`], *t* = `r round(exp2BHNBeta$statistic[4], digits = 2)`, *p* `r format.pval(exp2BHNBeta$p.value[4], digits = 2, eps = 0.001, nsmall = 3)`. As can be seen in Figure \@ref(fig:figure2), simple slopes showed that the more participants attributed socially desirable human traits to criminals, the less harshly participants thought they should be punished, *b* = `r round(exp2BHNSlope$simslopes$slopes$Est.[1], digits = 2)`, [`r round(exp2BHNSlope$simslopes$slopes$"2.5%"[1], digits = 2)`, `r round(exp2BHNSlope$simslopes$slopes$"97.5%"[1], digits = 2)`], *t* = `r round(exp2BHNSlope$simslopes$slopes$"t val."[1], digits = 2)`, *p* = `r format.pval(exp2BHNSlope$simslopes$slopes$"p"[1], digits = 1)`. The more participants attributed socially undesirable human traits to criminals, the more harshly participants thought they should be punished, *b* = `r round(exp2BHNSlope$simslopes$slopes$Est.[2], digits = 2)`, [`r round(exp2BHNSlope$simslopes$slopes$"2.5%"[2], digits = 2)`, `r round(exp2BHNSlope$simslopes$slopes$"97.5%"[2], digits = 2)`], *t* = `r round(exp2BHNSlope$simslopes$slopes$"t val."[2], digits = 2)`, *p* `r format.pval(exp2BHNSlope$simslopes$slopes$"p"[2], digits = 2, eps = 0.001, nsmall = 3)`. The model explained approximately `r round((exp2BHNR2$r.squared * 100), digits = 0)`% of the variance in the harshness of punishment scores, $R^2$ = `r round(exp2BHNR2$r.squared, digits = 2)`, $F$(`r round(exp2BHNR2$fstatistic[2], digits = 0)`, `r round(exp2BHNR2$fstatistic[3], digits = 0)`) = `r round(exp2BHNR2$fstatistic[1], digits = 2)`. These data suggest that the apparent relationships between animalistic and mechanistic dehumanization and punishment reported in previous research [@Bastian2013] are better explained by the social desirability of the traits.

@@

In study 3A, we used an experimental design to examine further the hypothesized causal relationship between trait-based dehumanization and punishment when controlling for the social desirability of human traits incorporated into the stimuli. We described criminals with traits that varied in desirability and perceived humanness creating a $2 \times 2$ design. We then measured participants' willingness to endorse parole for each criminal described. The dual model predicts that criminals who are described in uniquely human terms will be more likely to be granted parole. We predicted that criminals described in socially desirable terms would be more likely to be granted parole. In principle, this design allows us to detect independent effects of dehumanization and trait sociability or an interaction between the two. In study 3A, we specifically measure the extent to which animalistic dehumanization is causally related to parole decisions. Thus, our measures included uniquely human traits and those shared with other animals. We predict that participants will be more likely to endorse parole for criminals described with socially desirable traits, regardless of whether or not those traits are uniquely human or shared with other animals.

@@

A power analysis using G\*Power, with effect size specification as in SPSS, indicated that a sample size of 135 would allow us to detect a medium effect size ($\eta_p^2$ = .09) with a $2 \times 2$ factorial, repeated measures design, an alpha of .05, and a power of .95. To counterbalance the sample equally, a sample of 136 participants was collected, of whom `r gen_data3A[1]` identified as female, `r gen_data3A[2]` as male, `r gen_data3A[3]` as non-binary, and `r gen_data3A[4]` who preferred not to indicate their gender identity.  Ages ranged from `r min_age3A` to `r max_age3A` (*M* = `r round(age_mean3A, digits = 1)`, *SD* = `r round(age_sd3A, digits = 2)`).  All participants were adults fluent in English who had never been to prison for committing a crime.  Owing to a noticeable increase in failed attention checks during pilot data collection, the minimal approval rating on Prolific was raised from 90 to 95%.  Despite this, data from 46 participants were omitted and replaced due to failed attention checks^[The high rate of failed attention checks was likely owing to a wider issue with Prolific at the time of data collection, a surge in new users after a social media post promoting the platform went viral [@Charalambides2021]]. Three participants were mistakenly recruited after the intended sample size had been met, and thus, their data were excluded from analyses. Including data submitted by excess participants in analyses yielded the same results as those reported. Participation took an average of seven minutes.

@@

**Vignettes**. All participants responded to the same four vignettes, each detailing a different scenario in which a criminal's eligibility for parole was assessed. Efforts were made to ensure that all four vignettes were similar in length, degree of detail, and contextual aspects, such as how long the criminal had spent in prison and who was described as attributing the traits to the criminal. In each vignette, the criminal's name and pronouns were gender-neutral, their age and ethnicity were not indicated, and their crime and sentence were not specified. The four vignettes for study 3A can be seen in the electronic supplementary materials. In the uniquely human socially desirable condition, the criminal was described as *cultured, civilized, sophisticated*, and *moral*, while in the uniquely human condition socially undesirable, the criminal was described as *corrupt, controlling, arrogant* and *bitter*. In the animalistic desirable condition, the criminal was described as *energetic, trusting, genuine* and *having curiosity*, while in the animalistic undesirable condition, the criminal was described as *uncultured, unrefined, unsophisticated*, and *stupid*.

@@

**Agreement with parole**. The dependent variable, agreement with granting parole, was measured using the following single-item measure: '*I think (Alex/Sam/Robin/Jamie) should be granted parole*'. This measure appeared after each vignette, and participants indicated their agreement using an unmarked sliding scale ranging from Strongly Disagree (0) to Strongly Agree (100). The slider's starting point was always centred at 50.

@@

```{r figure3, fig.cap="Results of studies 3A and 3B.  Criminals described with undesirable traits were less likely to be granted parole than were those described with desirable traits, regardless of whether or not those traits were uniquely human.", fig.align="center", out.width='90%', echo=FALSE}

@@

#calculate participant mean/sd for ages

@@

The agreement with granting parole scale and attention check were the same as those used in study 3A.
